Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-26 LICS21/SPCF.tex
Mainly LaTeX reformatting to fit 2-column style
\appendices

\section{Statistical PCF}
\label{sec:SPCF}

@@

The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
\begin{equation}
\red(M,s) = \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation}
\lo{Both $\to$ and $\red$ are called reduction relation.}

The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value. A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.

@@

\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}

For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

\paragraph{}
Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.

\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

>>>>

% !TEX root = main.tex
\section{Statistical PCF}
\label{sec:SPCF}

@@

The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
\begin{equation*}
\red(M,s) = \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N\\
        & \qquad \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation*}
\lo{Both $\to$ and $\red$ are called reduction relation.}

The result after $n$ steps is then simply 
\[
\red^n(M,s) = \underbrace{\red(\ldots\red(}_n M,s)\ldots)
\] 
and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value. A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.

@@

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

%\paragraph{}
Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.

%\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

### 2020-12-26 LICS21/supermartingales.tex
\appendices

\begin{definition}\rm
A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$

    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
\end{enumerate}
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
\end{definition}

\paragraph{}

@@

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Theorem 7.2]{DBLP:conf/popl/FioritiH15}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \dfrac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

\begin{proof}
We first prove by induction on $n$: 
\[
\forall n \in \omega \, . \, \expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \Big(\sum_{i=0}^{n-1} \mathbb{P}[T > i]\Big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}. 
In Lemma 5.5 of op.~cit., the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}

@@

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
\[
s \in U_i
\iff
(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
\;
\hbox{for some }r \in \Real^{j_i}
\]
(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)

Equivalently
\(
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\)
%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

@@

Take $s \in U_i$, and set $l = l_i$.
Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Take $u \in (U_i)_{\leq l}$.
It then follows from the definition of ranking function that
\[
\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\]
Take $A \in \calF_n$, and integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
\end{align*}
and so, by (\ref{eqn:truncate})
\begin{align}
& \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
& \leq 
\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
\label{eqn:a u ui}
\end{align}

>>>>

\begin{definition}\rm
A \emph{ranking function on $M \in \Lambda^0$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z)$ where $z$ is not free in $N$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$
    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
\end{enumerate}
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value; and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.

It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
\end{definition}

%\paragraph{}

@@

The following is an iterated version of Doob's well-known Optional Sampling Theorem (see, e.g., \cite[\S 6.7]{AshDD00} and \cite[Thm.~7.2]{DBLP:conf/popl/FioritiH15}).
\begin{theorem}[Optional Sampling]
\label{thm:optional sampling}
Let $(X_n)_{n\in \omega}$ be a supermartingale, and $(T_n)_{n \in \omega}$ a sequence of increasing stopping times, both adapted to filtration $(\calF_n)_{n \in \omega}$, then $(X_{T_n})_{n \in \omega}$ is a supermartingale adapted to $(\calF_{T_n})_{n \in \omega}$ if one of the following conditions holds:
\begin{enumerate}
\item each $T_n$ is bounded i.e.~$T_n < c_n$ where $c_n$ is a constant
\item $(X_n)_{n\in \omega}$ is uniformly integrable.
\end{enumerate}
\end{theorem}

\begin{lemma}%[{\citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15}}]
% \citep[Prop 1]{DBLP:conf/popl/ChatterjeeFNH16}}
\label{lem:rank-PAST}
Let $(Y_n)_{n \in \omega}$ be a $\epsilon$-ranking supermartingale with respect to the stopping time $T$, then $T < \infty$ almost surely, and $\expect{T} \leq \frac{\expect{Y_0}}{\epsilon}$.
\end{lemma}

\begin{proof}
We first prove (by induction on $n$): for all $n \in \omega$
\[
%\forall n \in \omega \, . \, 
\expect{Y_n} \leq \expect{Y_0} - \epsilon \cdot \big(\textstyle\sum_{i=0}^{n-1} \mathbb{P}[T > i]\big).
\]
It follows that $\sum_{i=0}^{\infty} \mathbb{P}[T > i]$ converges; hence we have $\lim_{i \to \infty} \mathbb{P}[T > i] = 0$ and so $\mathbb{P}[T < \infty] = 1$.
It then remains to observe: if $T < \infty$ a.s., then $\expect{T} = \sum^\infty_{i=0} \mathbb{P}[T > i]$.
\end{proof}

\changed[lo]{\Cref{lem:rank-PAST} is a slight extension of \citep[Lemma 5.5]{DBLP:conf/popl/FioritiH15},
wherein the stopping time $T$ is assumed to be the random variable $s \mapsto \min\set{n \mid Y_n(s) = 0}$.}

@@

\begin{proof}
We show the non-trivial case of $i = 2$.
First we express 
\begin{equation}
f(M_{n+1})[s \in \mathbf{T}_2] = \sum_{i \in \calI} 
f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])[s \in U_i]
\label{eqn:f 2 n+1}
\end{equation}
where 
\begin{itemize}
\item $\calI$ is a countable indexing set
\item $E_i[\cdot][\tsample] \in \mathsf{Sk}_{j_i}$, and $\sigma_i : S \to \Real^{j_i}$, and $\rho(s) := \pi_h(\pi_1(\red^n(M, s))) = \pi_h(\pi_t^{l_i}(s)) \in \Real$ 
\item $\set{U_i}_{i \in \calI}$ is a partition of $\mathbf{T}_2$, 
where each $U_i$ is determined by a skeletal term $E_i[\cdot][\tsample]$ and a number (of draws) $l_i \leq n$, satisfying
\(
s \in U_i
\iff
(E_i[\underline{r}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s),
\;
\hbox{for some }r \in \Real^{j_i}.
\)
(We use the (measurable) function $\sigma_i : S \to \Real^{j_i}$ to skolemise the existentially quantified $r$.)
Equivalently
\[
U_i := \#\mathrm{draw}_n^{-1}(l_i) \cap M_n^{-1}(E_i[\cdot][\tsample], \Real^{j_i}) \in \calF_n.
\]
%\lo{I.e.~$(E_i[\underline{\sigma_i(s)}][{\tsample}], \pi_t^{l_i}(s)) = \red^n(M, s)$ iff $s \in U_i$.}
\end{itemize}

@@

Take $s \in U_i$, and set $l = l_i$.
Plainly $\sigma_i(s)$ depends on $s_{\leq l}$, and $\rho(s)$ depends on $s_{\leq l +1}$.
Take $u \in (U_i)_{\leq l}$.
It then follows from the definition of ranking function that
\[%\textstyle
\int_I \Leb(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \leq f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\]
Take $A \in \calF_n$, and integrating both sides, we get
\begin{align*}
& \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, \int_I \textrm{Leb}(\dif r) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\underline{r}])) \\
& \leq \int_{(A \cap U_i)_{\leq l}} \Leb_l(\dif u) \, f(E_i[\underline{\widehat{\sigma_i}(u)}][\tsample])).
\end{align*}
Since $\Leb_{l+1}$ is the (unique) product measure satisfying $\Leb_{l+1}(V \times B) = \Leb_l(V) \cdot \Leb(B)$, and $(U_i)_{\leq l_i} \cdot I^\omega = U_i$, we have
%it follows from the preceding observation that
\begin{align*}
& \int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\underline{\widehat{\rho}(u')}])) \\
& \leq 
\int_{(A \cap U_i)_{\leq l+1}} \Leb_{l+1}(\dif u') \, f(E_i[\underline{\widehat{\sigma_i}(u')}][\tsample]))
\end{align*}
and so, by (\ref{eqn:truncate})
\begin{align}
& \int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\underline{\rho(s)}])) \nonumber \\
& \leq 
\int_{A \cap U_i} \mu(\dif s) \, f(E_i[\underline{\sigma_i(s)}][\tsample])).
\label{eqn:a u ui}
\end{align}
